One-armed spirals in locally isothermal, radially structured self-gravitating discs

### 2014-11-15 paper/paper.tex
minor updates

\title[One-arm spirals]{One-arm spirals in 
  locally isothermal, radially structured self-gravitating discs}

@@

\maketitle
\begin{abstract} 
  We present numerical simulations of astrophysical discs with an 
  imposed temperature profile and in which a sub-portion of it is
  self-gravitating. This might represent, for example, mass  
  built-up in a circumstellar or protoplanetary
  disc at the outer edge of a `dead zone', in which there
  is inefficient magneto-hydrodynamic turbulence to enable accretion. 
  We conduct numerical gravito-hydrodynamic experiments in 2D using
  the FARGO code and in 3D with the ZEUS-MP and PLUTO codes. In such
  discs we observe the development of a 
  spiral mode with azimuthal wavenumber $m=1$ and co-rotation radius 
  outside the bulk of the spiral arm, resulting in a nearly stationary 
  one-arm spiral pattern in the self-gravitating region of the
  disc. The development of this one-arm 
  spiral does not require the movement of  
  the central star, as found in previous studies. Instead, we find it 
  is associated with the imposed 
  temperature gradient and the radial disc structure. 
  We apply results from linear theory and attribute the
  growth of this spiral to the non-conservation of angular momentum
  for non-axisymmetric linear perturbations in discs with a forced temperature
  gradient. This may lead to an unstable exchange of angular momentum between the spiral 
  and the background disc. However, because the interaction does
  not explicitly depend on self-gravity, we suggest it may also
  affect low-frequency one-arm spirals in non-self-gravitating
  discs. 
\end{abstract}

@@

% original motivation for this work (DZGI) 
A wide range of GIs can occur in  gaseous 
astrophysical discs 
\citep{papaloizou89,christo92,christo93,hadley11,hadley14}. It is then
of theoretical importance to clarify what type of GI is relevant to 
partially self-gravitating discs, such as a protoplanetary disc containing a 
massive dead zone.  %This was indeed the original goal of this project. 
This work presents a first step towards this goal. Specifically, we 
study a class of seemingly simple disc models which were found to 
consistently produce \emph{one-armed} spiral structures, despite not 
satisfying the usual requirements to generate them.

@@

%brief history of m=1 
%stellar wobble from indiana group  
Single-arm spirals, or eccentric modes, corresponding to perturbations 
with azimuthal wavenumber $m=1$, have received interest in
the context of protoplanetary discs because of their global nature
\citep{adams89,heemskerk92,laughlin96,tremaine01,papaloizou02,hopkins10}. 
In the `SLING' mechanism proposed by \cite{shu90}, the $m=1$ spiral
instability is attributed to motion of the central star induced by the
one-arm perturbation and requires a massive disc \citep[the former may
have observable consequences, ][]{michael10}. However, our numerical
models do not have these features.

@@

The aforementioned analytical studies have adopted barotropic, smooth disc models. 
In this work, we identify a mechanism to generate one-arm
spiral patterns that is related to a non-barotropic equation of
state and disc structure. We find that if the disc temperature or
sound-speed is a prescribed function of position, then 
one-arm spiral modes can grow inside a cavity that results from the
adopted disc structure.

Our study is based on direct hydrodynamic simulations using three
different grid-based codes, both in two-dimensions (2D) and
three-dimensions (3D). We also apply results from linear
theory to explain their growth. We find that, when the disc
temperature is prescribed, the usual statement for the conservation of
angular momentum for linear perturbations acquires a source term
proportional to the temperature gradient, which may act to destabilise
$m=1$ perturbations. Although we work with self-gravitating discs,
this source term does not depend explicitly on disc self-gravity. Thus, the
destabilisation effect we describe should also be applicable to
non-self-gravitating discs.

>>>>

\title[One-armed spirals]{One-armed spirals in 
  locally isothermal, radially structured self-gravitating discs}

@@

\maketitle
\begin{abstract} 
  We present numerical simulations of astrophysical discs with an 
  imposed temperature profile and in which a sub-portion of it is
  self-gravitating. This might represent, for example, mass  
  built-up in a circumstellar or protoplanetary
  disc at the outer edge of a `dead zone', in which there
  is inefficient magneto-hydrodynamic turbulence to enable accretion. 
  We conduct numerical gravito-hydrodynamic experiments in 2D using
  the FARGO code and in 3D with the ZEUS-MP and PLUTO codes. In such
  discs we observe the development of a 
  spiral mode with azimuthal wavenumber $m=1$ and co-rotation radius 
  outside the bulk of the spiral arm, resulting in a nearly stationary 
  one-armed spiral pattern in the self-gravitating region of the
  disc. The development of this one-armed 
  spiral does not require the movement of  
  the central star, as found in previous studies. Instead, we find it 
  is associated with the imposed 
  temperature gradient and the radial disc structure. 
  We apply results from linear theory and attribute the
  growth of this spiral to the non-conservation of angular momentum
  for non-axisymmetric linear perturbations in discs with a forced temperature
  gradient. This may lead to an unstable exchange of angular momentum between the spiral 
  and the background disc. However, because the interaction does
  not explicitly depend on self-gravity, we suggest it may also
  affect low-frequency one-armed spirals in non-self-gravitating
  discs. 
\end{abstract}

@@

% original motivation for this work (DZGI) 
A wide range of GIs can occur in  gaseous 
astrophysical discs 
\citep{papaloizou89,christo92,christo93,hadley11,hadley14}. It is then
of theoretical importance to clarify what type of GI is relevant to 
partially self-gravitating discs, such as a protoplanetary disc containing a 
massive dead zone.  %This was indeed the original goal of this project. 
This work presents a first step towards this goal. Specifically, here we 
study a class of seemingly simple disc models which were found to 
consistently produce \emph{one-armed} spiral structures, despite not 
satisfying the usual requirements to generate them.

@@

%brief history of m=1 
%stellar wobble from indiana group  
Single-arm spirals, or eccentric modes, corresponding to perturbations 
with azimuthal wavenumber $m=1$, have received interest in
the context of protoplanetary discs because of their global nature
\citep{adams89,heemskerk92,laughlin96,tremaine01,papaloizou02,hopkins10}. 
In the `SLING' mechanism proposed by \cite{shu90}, the $m=1$ spiral
instability is attributed to motion of the central star induced by the
one-armed perturbation and requires a massive disc \citep[the former may
have observable consequences, ][]{michael10}. However, our numerical
models do not have these features.

@@

The aforementioned analytical studies have adopted barotropic, smooth disc models. 
In this work, we identify a mechanism to generate one-armed
spiral patterns that is related to a non-barotropic equation of
state and disc structure. We find that if the disc temperature or
sound-speed is a prescribed function of position, then 
one-armed spiral modes can grow inside a cavity that results from the
adopted disc structure.

Our study is based on direct hydrodynamic simulations using three
different grid-based codes, both in two-dimensions (2D) and
three-dimensions (3D). We also apply linear
theory to interpret simulation results. We find that, when the disc
temperature is prescribed, the usual statement for the conservation of
angular momentum for linear perturbations acquires a source term
proportional to the temperature gradient, which may act to destabilise
$m=1$ perturbations. Although we work with self-gravitating discs,
this source term does not depend explicitly on disc self-gravity. Thus, the
destabilisation effect we describe should also be applicable to
non-self-gravitating discs.

### 2014-11-15 paper/results.tex
minor updates

Fig. \ref{2d_angmom} shows the evolution of disc angular momentum
components. Only the $m=0,\,1$ components are 
plotted since they are dominant. The $m=1$ structure has
an associated negative angular momentum.  
Its growth is compensated by an increase in the axisymmetric
component of angular momentum, such that $\Delta J_0 + \Delta
J_1 \sim 0$. Note that FARGO does not conserve angular momentum
exactly. However, we find the total angular momentum varies by 
$|\Delta J/J|= O(10^{-6})$, and is much smaller than the
change in the angular momenta components, $|\Delta J_{0,1}/J|>
O(10^{-5})$. Fig. \ref{2d_angmom} then suggest that angular momentum
is transferred from the one-arm spiral to the background disc.

@@

Fig. \ref{2d_fargo_viz} shows the $m=1$ surface density
at the end of this simulation. For this one-armed spiral
we measure a co-rotation radius and growth rate  
\begin{align*}
  &R_c \simeq 4.4R_0,\\
  &\gamma\simeq 0.014\Omega_k(R_0) = 0.13\Omega_p. 
\end{align*}
The spiral can be considered as low frequency because its 
pattern speed $\Omega_p \simeq 0.1\Omega_k(R_0)\lesssim 0.3\Omega$ in
$R\in[R_{1},R_{2}]$ (where it has the largest amplitude). The growth rate $\gamma$ is also slow
relative to the local rotation. (Although the characteristic growth
time $\gamma^{-1} \simeq 10P_0$.) Thus, the spiral 
pattern is nearly stationary.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/polarxy2_dens120_fargo.ps}
  \caption{Cartesian visualisation of the $m=1$ surface density
    structure in the FARGO simulation initialised with only $m=1$
    perturbations. 
    \label{2d_fargo_viz}} 
\end{figure}

@@

Using the estimated value of $R_c$, we plot in
Fig. \ref{fargo_qbarrier} the quantity $\nu^2 - 1 + Q^{-2}$, which is
required to be positive in local theory for purely wave-like
solutions to the dispersion relation (Eq. \ref{dispersion}) when the
mode frequency is given. %Note that, with the above measured wavenumber $k$, 
%Eq. \ref{dispersion} requires $\nu^2 - 1 + Q^{-2}\sim0$ and the lower
%sign (long waves) to be taken. 
Fig. \ref{fargo_qbarrier} shows two $Q$-barriers located in
the inner disc, at $R_{Qb}=R_0$ and $R_{Qb}=1.6R_0$; the bounded region is indeed the
where the $m=1$ spiral develops. This suggests that the one-arm 
spiral is trapped. Note in this region, $\nu^2 - 1 + 
Q^{-2}\simeq 0.1\ll 1$, as required for consistency with
Eq. \ref{dispersion} with the lower sign taken (i.e. long waves).  
There is one outer Lindblad resonance at $R_L\simeq
7.2R_0$. Thus, acoustic waves may be launched in $R\gtrsim 7.2R_0$ by
the spiral disturbance in the inner disc \citep{lin11b}. 
%implications: effect on dynamics in outer disc

@@

\subsection{Angular momentum exchange with the background disc}  
The wavenumber $k = \pi G\Sigma/c_s^2$ minimises the shifted frequency
$\sbar$ in the dispersion relation, Eq. \ref{dispersion}.  
%\emph{axisymmetric} perturbations. 
Physically, then, it is not
surprising to find perturbations of this wavenumber in a
self-gravitating disc where $Q\sim 1$. However, according the
local dispersion relation, low-frequency $m=1$ perturbations are
formally stable for real $k$.

@@

Fig. \ref{fargo_varq} compares the $m=1$ spiral amplitudes as a
function of $q$. We indeed observe slower growth with decreasing
$q$. Although the figure indicates growth for the strictly isothermal
disc ($q=0$), we did not observe a coherent one-arm spiral upon
inspection of the $m=1$ surface density field. The growth in this case
may be associated with higher $m$ modes, which dominated the
simulation.    
% show result of isothermal calc?

@@

We plot growth rates of the $m=1$ spiral as a function of $q$ in 
Fig. \ref{fargo_varq_growth}. The correlation can be fitted with a
linear relation
\begin{align*}
  \gamma \simeq \left[0.015 q - 7.9\times10^{-4}\right] \Omega_k(R_0). 
\end{align*}
% difficult to measure very small growth rates
This confirms that the temperature gradient is responsible for the
development of the one-arm spirals observed in our simulations. 
\begin{figure}
  \includegraphics[width=\linewidth]{figures/m1_analysis_plot_ratemax_fargo_varq.ps}    
  \caption{Growth rates of the $m=1$ spiral mode as a function of the
    imposed sound-speed gradient $q$ (asterisks). A linear fit is also
    plotted (dotted line). 
    \label{fargo_varq_growth}} 
\end{figure}
%maybe extend the low q simulations to get better rates?

>>>>

Fig. \ref{2d_angmom} shows the evolution of disc angular momentum
components. Only the $m=0,\,1$ components are 
plotted since they are dominant. The $m=1$ structure has
an associated negative angular momentum.  
Its growth is compensated by an increase in the axisymmetric
component of angular momentum, such that $\Delta J_0 + \Delta
J_1 \sim 0$. Note that FARGO does not conserve angular momentum
exactly. However, we find the total angular momentum varies by 
$|\Delta J/J|= O(10^{-6})$, and is much smaller than the
change in the angular momenta components, $|\Delta J_{0,1}/J|>
O(10^{-5})$. Fig. \ref{2d_angmom} then suggest that angular momentum
is transferred from the one-armed spiral to the background disc.

@@

Fig. \ref{2d_fargo_viz} shows the $m=1$ surface density
at the end of this simulation. For this one-armeded spiral
we measure a co-rotation radius and growth rate  
\begin{align*}
  &R_c \simeq 4.4R_0,\\
  &\gamma\simeq 0.014\Omega_k(R_0) = 0.13\Omega_p. 
\end{align*}
The spiral can be considered as low frequency because its 
pattern speed $\Omega_p \simeq 0.1\Omega_k(R_0)\lesssim 0.3\Omega$ in
$R\in[R_{1},R_{2}]$ (where it has the largest amplitude). Thus, the spiral 
pattern appears nearly stationary. The growth rate $\gamma$ is also slow
relative to the local rotation, although the characteristic growth
time $\gamma^{-1} \simeq 10P_0$ is not very long.  
\begin{figure}
  \includegraphics[width=\linewidth]{figures/polarxy2_dens120_fargo.ps}
  \caption{Cartesian visualisation of the $m=1$ surface density
    structure in the FARGO simulation initialised with only $m=1$
    perturbations. 
    \label{2d_fargo_viz}} 
\end{figure}

@@

Using the estimated value of $R_c$, we plot in
Fig. \ref{fargo_qbarrier} the quantity $\nu^2 - 1 + Q^{-2}$, which is
required to be positive in local theory for purely wave-like
solutions to the dispersion relation (Eq. \ref{dispersion}) when the
mode frequency is given. %Note that, with the above measured wavenumber $k$, 
%Eq. \ref{dispersion} requires $\nu^2 - 1 + Q^{-2}\sim0$ and the lower
%sign (long waves) to be taken. 
Fig. \ref{fargo_qbarrier} shows two $Q$-barriers located in
the inner disc, at $R_{Qb}=R_0$ and $R_{Qb}=1.6R_0$; the bounded region is indeed the
where the $m=1$ spiral develops. This suggests that the one-armed 
spiral is trapped. Note in this region, $\nu^2 - 1 + 
Q^{-2}\simeq 0.1\ll 1$, which is necessary for consistency with 
the measured wavenumber $k$ and Eq. \ref{wavenumber}.  
%Eq. \ref{dispersion} with the lower sign taken (i.e. long waves).  
There is one outer Lindblad resonance at $R_L\simeq
7.2R_0$. Thus, acoustic waves may be launched in $R\gtrsim 7.2R_0$ by
the spiral disturbance in the inner disc \citep{lin11b}. 
%implications: effect on dynamics in outer disc

@@

\subsection{Angular momentum exchange with the background disc}  
The wavenumber $k = \pi G\Sigma/c_s^2$ minimises the shifted frequency
$\sbar$ in the dispersion relation, Eq. \ref{dispersion}.  
%\emph{axisymmetric} perturbations. 
Physically, then, it is not
surprising to find perturbations of this wavenumber in a
self-gravitating disc where $Q\sim 1$. However, according the
local dispersion relation, $m=1$ perturbations are
formally stable for real $k$.

@@

Fig. \ref{fargo_varq} compares the $m=1$ spiral amplitudes as a
function of $q$. We indeed observe slower growth with decreasing
$q$. Although the figure indicates growth for the strictly isothermal
disc ($q=0$), we did not observe a coherent one-armed spiral upon
inspection of the $m=1$ surface density field. The growth in this case
may be associated with higher $m$ modes, which dominated the
simulation.    
% show result of isothermal calc?

@@

We plot growth rates of the $m=1$ spiral as a function of $q$ in 
Fig. \ref{fargo_varq_growth}. The correlation can be fitted with a
linear relation
\begin{align*}
  \gamma \simeq \left[0.015 q - 7.9\times10^{-4}\right] \Omega_k(R_0). 
\end{align*}
% difficult to measure very small growth rates
This confirms that the temperature gradient is responsible for the
development of the one-armed spirals observed in our simulations. 
\begin{figure}
  \includegraphics[width=\linewidth]{figures/m1_analysis_plot_ratemax_fargo_varq.ps}    
  \caption{Growth rates of the $m=1$ spiral mode as a function of the
    imposed sound-speed gradient $q$ (asterisks). A linear fit is also
    plotted (dotted line). 
    \label{fargo_varq_growth}} 
\end{figure}
%maybe extend the low q simulations to get better rates?

### 2014-11-15 paper/results_3d.tex
minor updates

Visualisations of the 3D simluations are shown in   
Fig. \ref{3d_prelim} for the disc midplane and near the upper disc
boundary. The snapshots are chosen when the one-armed spirals in the two codes
have reached similar amplitudes. The agreement between the codes,
as well as with the 2D simulations, is satisfactory. Away from the
disc midplane, a coherent spiral arm is not prominent, but the
disturbance is radially more global, extending up to $R\sim 4R_0$.   
%implications?

@@

\subsection{Vertical structure}
Fig. \ref{3d_rz} shows the vertical structure of the one-arm
spiral in the PLUTO run. The main spiral is confined to $z < H$. Thus,
a 2D disc model, representing dynamics near the disc midplane, 
should be sufficient to study the instability. Interestingly, however, for $R>2R_0$ the
spirals  increase in amplitude away from the midplane. It remains
small in our disc model ($|\Delta\rho| < 0.1$), but could become 
significant with a larger vertical domain. The figure shows that 
the one-arm spiral can have a global influence away from the
midplane of the self-gravitating annulus.

@@

%Vertical velocities
%do increase with the growth of the one-arm spiral, 
%Larger amplitudes
%are reached in ZEUS because of larger numerical noise. 
%but we find  $\Theta \lesssim 10^{-2}$.

@@

\begin{figure}
%scale=0.41
  \includegraphics[scale=.41,clip=true,trim=0cm 1cm 0cm 0cm]{figures/nonaxi_evol_ang_zeus}
  \includegraphics[scale=.41]{figures/nonaxi_evol_ang_pluto}
  \caption{Evolution of angular momentum components in the 3D 
    simulations. The perturbation 
    relative to $t=10P_0$, during the growth of the one-arm spiral,
    is shown in units of the initial total angular momentum
    $J_\mathrm{ref}$.\label{3d_angmom}}  
\end{figure}

>>>>

Visualisations of the 3D simluations are shown in   
Fig. \ref{3d_prelim} for the disc midplane and near the upper disc
boundary. The snapshots are chosen when the one-armeded spirals in the two codes
have reached similar amplitudes. The agreement between the codes,
as well as with the 2D simulations, is satisfactory. Away from the
disc midplane, a coherent spiral arm is not prominent, but the
disturbance is radially more global, extending up to $R\sim 4R_0$.   
%implications?

@@

\subsection{Vertical structure}
Fig. \ref{3d_rz} shows the vertical structure of the one-armed
spiral in the PLUTO run. The main spiral is confined to $z < H$. Thus,
a 2D disc model, representing dynamics near the disc midplane, 
should be sufficient to study the instability. Interestingly, however, for $R>2R_0$ the
spirals  increase in amplitude away from the midplane. It remains
small in our disc model ($|\Delta\rho| < 0.1$), but could become 
significant with a larger vertical domain. The figure shows that 
the one-armed spiral can have a global influence away from the
midplane of the self-gravitating annulus.

@@

%Vertical velocities
%do increase with the growth of the one-armed spiral, 
%Larger amplitudes
%are reached in ZEUS because of larger numerical noise. 
%but we find  $\Theta \lesssim 10^{-2}$.

@@

\begin{figure}
%scale=0.41
  \includegraphics[scale=.41,clip=true,trim=0cm 1cm 0cm 0cm]{figures/nonaxi_evol_ang_zeus}
  \includegraphics[scale=.41]{figures/nonaxi_evol_ang_pluto}
  \caption{Evolution of angular momentum components in the 3D 
    simulations. The perturbation 
    relative to $t=10P_0$, during the growth of the one-armed spiral,
    is shown in units of the initial total angular momentum
    $J_\mathrm{ref}$.\label{3d_angmom}}  
\end{figure}

### 2014-11-15 paper/summary.tex
minor updates

\section{Summary and conclusions}\label{summary}
In this paper, we have carried out direct numerical hydrodynamic
simulations of radially structured, self-gravitating locally
isothermal discs.     
We find such systems can be unstable to low-frequency perturbations
with azimuthal wavenumber $m=1$, which lead to the development of one-arm 
trailing spirals that persist for at least $O(10^2)$ orbits. The 
spiral pattern speed is smaller than the local disc rotation, making
it almost stationary. The growth rates are $O(10^{-2}\Omega)$ which
gives a characteristic growth time of $O(10P_0)$. 
%While this is slow
%compared to the local rotation, it is dynamical at co-rotation.

We used three independent numerical codes --- FARGO in 2D, ZEUS-MP and
PLUTO in 3D --- to verify the growth of  
one-arm spirals in our disc models is due to the imposed temperature
profile and disc structure. The spiral is confined between two
$Q$-barriers owing to a surface density bump. Growth rates increased
with the magnitude of the imposed temperature gradient, and one-armed
spirals did not develop in strictly isothermal simulations.  
We find no significant vertical 
motion in 3D simulations, which indicates the instability is
two-dimensional.

We applied angular momentum conservation within linear theory to
explain our numerical results. A  
forced temperature gradient introduces a torque on linear 
perturbations. We called this the background torque because it
represents an exchange of angular momentum between the background disc
and the perturbations. In the local approximation, we showed that this
background torque is negative for trailing spirals, which enforces  
low-frequency $m=1$ modes because they are associated with negative
angular momentum. We therefore interpret the one-arm spiral instability
observed in our simulations as a neutral, tightly-wound $m=1$ mode
being destabilised by the imposed temperature gradient.

%works in 2D/3D
%forced gradient feedback 
%neutral mode destabilized      
\subsection{Spectulations and future work}
%balancing heating and cooling 
In the deeply non-linear regime, the one-arm spiral may 
shock and dissipate, depositing negative angular momentum onto
the background disc. The spiral amplitude would saturate by gaining
positive angular momentum. However, if the temperature gradient is
forced, it may be possible to achieve a balance between the gain of negative
angular momentum through the forced temperature gradient, and the gain
of positive angular momentum through shock dissipation. In order to
study this possibility, improved numerical models are needed to ensure 
total angular momentum conservation in long-term simulations.

%numerical simulations to include cooling/heating
A locally isothermal disc represents the limit of infinitely short
cooling (and heating) timescales, so the disc temperature instantly
returns to its inital value. %so we don't go to deeply non-linear
                             %regime 
This assumption can be relaxed by including an energy
equation with thermal relaxation. Preliminary FARGO simulations
indicate a relaxation timescale $t_c < 0.1\Omega_k^{-1}$ is needed for
the one-arm spiral to develop. This will be presented in a follow-up
study. %prelim sims show they are long lasting

%non-sg sims, Be stars, eccentric discs
% We stress that destabilisation through the temperature gradient
% does not explicitly depend on self-gravity. In principle, this effect
% will destabilise low-frequency $m=1$ trailing spirals regardless of
% its origin.
In our models and interpretation, self-gravity coupled with the disc 
structure permits a neutral one-arm mode, which is then destabilised.   
It would be interesting to explore other origins of the neutral
one-arm spiral. One possibility is in Be star discs
\citep{rivinius13}, for which 
one-arm oscillations may explain long-timescale variations in their
emission lines \citep[see e.g.][and references
therein]{okasaki97,papaloizou06c,ogilvie08}. These studies employ
strictly isothermal disc models, so it would be interesting to explore
the effect of a temperature gradient on the stability of these
oscillations.

>>>>

\section{Summary and conclusions}\label{summary}
In this paper, we have carried out direct numerical hydrodynamic
simulations of radially structured, self-gravitating locally
isothermal discs.     
We find such systems can be unstable to low-frequency perturbations
with azimuthal wavenumber $m=1$, which lead to the development of one-armed 
trailing spirals that persist for at least $O(10^2)$ orbits. The 
spiral pattern speed is smaller than the local disc rotation, making
it almost stationary. The growth rates are $O(10^{-2}\Omega)$ which
gives a characteristic growth time of $O(10P_0)$. 
%While this is slow
%compared to the local rotation, it is dynamical at co-rotation.

We used three independent numerical codes --- FARGO in 2D, ZEUS-MP and
PLUTO in 3D --- to verify the growth of  
one-armed spirals in our disc models is due to the imposed temperature
profile and disc structure. The spiral is confined between two
$Q$-barriers owing to a surface density bump. Growth rates increased
with the magnitude of the imposed temperature gradient, and one-armeded
spirals did not develop in strictly isothermal simulations.  
We find no significant vertical 
motion in 3D simulations, which indicates the instability is
two-dimensional.

We applied angular momentum conservation within linear theory to
explain our numerical results. A  
forced temperature gradient introduces a torque on linear 
perturbations. We called this the background torque because it
represents an exchange of angular momentum between the background disc
and the perturbations. In the local approximation, we showed that this
background torque is negative for trailing spirals, which enforces  
low-frequency $m=1$ modes because they are associated with negative
angular momentum. We therefore interpret the one-armed spiral instability
observed in our simulations as a neutral, tightly-wound $m=1$ mode
being destabilised by the imposed temperature gradient.

%works in 2D/3D
%forced gradient feedback 
%neutral mode destabilized      
\subsection{Spectulations and future work}
%balancing heating and cooling 
In the deeply non-linear regime, the one-armed spiral may 
shock and dissipate, depositing negative angular momentum onto
the background disc. The spiral amplitude would saturate by gaining
positive angular momentum. However, if the temperature gradient is
forced, it may be possible to achieve a balance between the gain of negative
angular momentum through the forced temperature gradient, and the gain
of positive angular momentum through shock dissipation. In order to
study this possibility, improved numerical models are needed to ensure 
total angular momentum conservation in long-term simulations.

%numerical simulations to include cooling/heating
A locally isothermal disc represents the limit of infinitely short
cooling (and heating) timescales, so the disc temperature instantly
returns to its inital value. %so we don't go to deeply non-linear
                             %regime 
This assumption can be relaxed by including an energy
equation with thermal relaxation. Preliminary FARGO simulations
indicate a relaxation timescale $t_c < 0.1\Omega_k^{-1}$ is needed for
the one-armed spiral to develop. This will be presented in a follow-up
study. %prelim sims show they are long lasting

%non-sg sims, Be stars, eccentric discs
% We stress that destabilisation through the temperature gradient
% does not explicitly depend on self-gravity. In principle, this effect
% will destabilise low-frequency $m=1$ trailing spirals regardless of
% its origin.
In our models and interpretation, self-gravity coupled with the disc 
structure permits a neutral one-armed mode, which is then
destabilised. Other origins of a neutral
one-armed spiral should be investigated. One possibility is in Be star discs
\citep{rivinius13}, for which 
one-armed oscillations may explain long-timescale variations in their
emission lines \citep[see e.g.][and references
therein]{okasaki97,papaloizou06c,ogilvie08}. These studies employ
strictly isothermal disc models, so it would be interesting to explore
the effect of a temperature gradient on the stability of these
oscillations.
